A theorical-practical approach to manifold learning

### 2018-11-05 DimReduction/Mds/Documentation/Article_mds.tex
Adding result section to Mds documentation

    As mentioned before, one of the goals of this work is to create a theorical
    development of the multidimentional scaling method. This section shows a
    detail explanation of the therical aspects of the method taking \cite{proof}
    as a base point.

@@

    Since each entry of the matrix $D^x$ can be expressed as
    $d_{ij}^{(x)} = x_i^t x_j$, the problem can be rewritten

@@

    Now, coming back to the original problem this information can be used to
    simplify the objective function

@@

    which implies that $Y = \hat{\Lambda}^{1/2} V^t$.

@@

    This method gives a way to find a matrix $D$ that encodes the distances
    between points that are in a low dimensional space. Now a procedure that
    takes a distance matrix as input and finds a corresponding configuration
    of points is described. The distance matrix is defined as follows

@@

    The unknown of this problem is a matrix $X$ that cotains the position of the
    $n$ points in the $d$ dimensional space.

@@

    the first step to solve the problem is rewrite each entry of matrix $D$ in
    a vector form

@@

    The goal is to express all the dot products of $x$ vectors in terms of the
    $d_{ij}^2$ distances. To achieve this goal, consider the following expression

@@

    But, using the restriction, one term can be canceled

@@

    Now in the other hand, the sum over $j$ would be

@@

    \begin{equation*}
        \begin{aligned}
            d_{ij}^2 &= x_i^t x_i - 2 x_i^t x_j + x_j^t x_j\\
            2 x_i^t x_j &= x_i^t x_i - d_{ij}^2 + x_j^t x_j\\
            x_i^t x_j &= \frac{1}{2} (x_i^t x_i - d_{ij}^2 + x_j^t x_j)\\
            \text{where}\\
            x_i^t x_i &= \frac{1}{n} \sum_{j=1}^n d_{ij}^2 - \frac{1}{n} \sum_{j=1}^n x_j^t x_j\\
            x_j^t x_j &= \frac{1}{n} \sum_{i=1}^n d_{ij}^2 - \frac{1}{n} \sum_{i=1}^n x_i^t x_i
        \end{aligned}
    \end{equation*}

    But this expression has a term that can be treated as duplicated, so the
    following propertie can be applied to simplify the expression

@@

    This is almost expressed in terms of the distances, the only problem is the
    last term. To express that part consider the following expression

@@

    Now the goal is completed, and the dot product between the $X$ components
    can be expressed in terms of distance operations

@@

    Which gives a way to compute all the points of the matrix $X$ finding a low
    dimensional representation of the original dataset.

@@

\begin{algorithm}
    $n$ = number of points\\
    $p$ = output dimensions\\
    $k$ = number of neighbors\\
    \vspace{0.5cm}
    $D_e$ = compute\_euclidean\_distance(data)\\
    $G$ = zeros($n$) //graph adjacency matrix\\
    \vspace{0.5cm}
    \For{i in 1:n}{
        $x_i$ = data($i$)\\
        $k_i$ = get\_nearest\_neighbors($x_i$, $k$)\\
        $G(i,k_i)$ = $D_e$($i$, $k_i$)\\
        $G(k_i,i)$ = $D_e$($i$, $k_i$)\\
    }
    \vspace{0.5cm}
    $H$ = $I - \frac{1}{n} e e^t$\\
    $D^x$ = compute\_distances($G$)\\
    $A$ = -$\frac{1}{2}$ pow\_elementwise($D^x$,2)\\
    $B$ = $HAH$\\
    $V$ = get\_eigenvectors($B$)\\
    $\Lambda$ = get\_eigenvalues($B$)\\
    $V$, $\Lambda$ = select\_smallers($p$, $V$, $\Lambda$)\\
    $Y$ = $\Lambda^{1/2} V^t$

@@

\cite{matlab}

@@

@@

    hence

>>>>

    As mentioned before, one of the goals of this work is to create a theorical
    development of the multidimentional scaling algorithm. This section shows a
    detail explanation of the therical aspects of the method taking \cite{proof}
    as base point.

@@

    Since each entry of the matrix $D^x$ can be expressed as
    $d_{ij}^{(x)} = x_i^t x_j$, then the problem can be rewritten as

@@

    Now, coming back to the original problem this information can be used to
    simplify the objective function as follows

@@

    which implies that $Y = \hat{\Lambda}^{1/2} V^t$. Which solves the problem.

@@

    This method gives a way to find a matrix $D^x$ that encodes the distances
    between points that are in a low dimensional space. Now, a procedure that
    takes a distance matrix as input and finds a corresponding configuration
    of points is described.

    First, the distance matrix is defined as follows

@@

    The unknown of this problem is a matrix $X$ that cotains the position of the
    $n$ points in a $d$ dimensional space.

@@

    To solve the problem the first step is rewrite each entry of matrix $D$ in
    a vector form

@@

    The goal is to express all the dot products of $x$ vectors in terms of the
    $d_{ij}^2$ distances. To achieve this goal, consider the following
    expressions

@@

    But, using the restriction, one term must be canceled

@@

    Now, in the other hand, the sum over $j$ would be

@@

    \begin{equation*}
        \begin{aligned}
            d_{ij}^2 &= x_i^t x_i - 2 x_i^t x_j + x_j^t x_j\\
            2 x_i^t x_j &= x_i^t x_i - d_{ij}^2 + x_j^t x_j\\
            x_i^t x_j &= \frac{1}{2} (x_i^t x_i - d_{ij}^2 + x_j^t x_j)\\
        \end{aligned}
    \end{equation*}

    Where the dot products of the left side can be replaced by the expressions
    shown below

    \begin{equation*}
        \begin{aligned}
            x_i^t x_i &= \frac{1}{n} \sum_{j=1}^n d_{ij}^2 - \frac{1}{n} \sum_{j=1}^n x_j^t x_j\\
            x_j^t x_j &= \frac{1}{n} \sum_{i=1}^n d_{ij}^2 - \frac{1}{n} \sum_{i=1}^n x_i^t x_i
        \end{aligned}
    \end{equation*}

    But this expression have a term that can be treated as duplicated, so the
    following propertie can be applied to simplify the expression

@@

    This is almost expressed in terms of distance operations, the only problem
    is the last term. To express that part consider the following expression

@@

    Now the goal is completed, and the dot product between the $x$ vectors
    can be expressed in terms of distance operations

@@

    Which gives a way to compute all the points of the matrix $X$, finding a low
    dimensional representation of the original dataset.

@@

In this section the implementation details of the method are exposed. The method
described in the previous section was developed using the Matlab programming
language \cite{matlab}. Algorithm \ref{alg1} shows a seudo-code of the
implemented method.

\begin{algorithm}
    $n$ = number of points\\
    $p$ = output dimensions\\
    $k$ = number of neighbors\\
    \vspace{0.5cm}
    $D_e$ = compute\_euclidean\_distance(data)\\
    $G$ = zeros($n$) //graph adjacency matrix\\
    \vspace{0.5cm}
    \For{i in 1:n}{
        $x_i$ = data($i$)\\
        $k_i$ = get\_nearest\_neighbors($x_i$, $k$)\\
        $G(i,k_i)$ = $D_e$($i$, $k_i$)\\
        $G(k_i,i)$ = $D_e$($i$, $k_i$)\\
    }
    $D^x$ = compute\_distances($G$)\\
    \vspace{0.5cm}
    $e$ = vector\_ones($n$)\\
    $H$ = $I - \frac{1}{n} e e^t$\\
    $A$ = -$\frac{1}{2}$ pow\_elementwise($D^x$,2)\\
    $B$ = $HAH$\\
    $V$ = get\_eigenvectors($B$)\\
    $\Lambda$ = get\_eigenvalues($B$)\\
    $V$, $\Lambda$ = select\_smallers($p$, $V$, $\Lambda$)\\
    $Y$ = $\Lambda^{1/2} V^t$

@@

This algorithm is divided in two main parts. The first one computes the initial
distance matrix, that represents the geodesic distance in the multidimensional
manifold. This computation is achieved by solving the shortest path problem
between all the pairs of nodes in a graph that represents the manifold. This
graph is constructed by taking each point as a node, and building an arc from
each node to its $k$ nearest neighbors. As a last step, the weight of each arc
is initialized as the euclidean distance between the two nodes connected.

And finally the procedure described in the theorical framework can be executed
to get a low dimensional representation of the original multidimensional
manifold.

@@

The method described in the previuos section was tested with 3 dimensional
manifolds in order to allow the visualization of the intial manifold and be able
to compare the quality of the resulting solution. Figure \ref{swiss} shows
a swiss roll dataset that was used as the multidimentional manifold. Figure
\ref{swiss_res} shows the result of the devolped method.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{images/swiss_roll.png}
    \caption{Swiss roll dataset}
    \label{swiss}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.7\linewidth]{images/swiss_roll_result.png}
    \caption{Swiss roll dataset reduced}
    \label{swiss_res}
\end{figure}

@@
